Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-01 tagless/tagless-final.tex
CBV CPS done

We can obtain CBV from CBN?

@@

>>>>

We can easily change our CBN CPS interpreter into a CBV CPS
interpreter, by replacing only one definition:
\begin{code}
module RCV = struct
  include RCN
  let lam f = {ko = 
    fun k -> k (fun e -> e.ko 
         (fun v -> f {ko = fun k -> k v}))}
end;;
\end{code}
Now when we apply an abstraction to an argument |e|, we always force
the evaluation of that argument before proceeding. This approach is in
line with Reynolds (cite) (albeit in a typed setting). The interpreter RCV is
useful for achieving the CBV evaluation of the object language
no matter whether the metalanguage is CBV or CBN.

In this section however we show a more general approach to CBV CPS:
CPS transformer. We take any implementation of Symantics (interpreter,
compiler, partial evaluator) and make it evaluate the CPS language
instead. We can view this transformation of interpreters as a
transformation on the object language, from direct to
continuation-passing style.

The transformation is canonical, as the following (abbreviated) code
shows
\begin{code}
module CPST(S: Symantics) = struct
  let int i = S.lam (fun k -> S.app k (S.int i))
  let add e1 e2 = S.lam (fun k ->
    S.app e1 (S.lam (fun v1 ->
    S.app e2 (S.lam (fun v2 -> S.app k (S.add v1 v2))))))
  let if_ ec et ef = S.lam (fun k ->
    S.app ec (S.lam (fun vc ->
    S.if_ vc (fun () -> S.app (et ()) k) (fun () -> S.app (ef ())
    k))))
  ...
  let lam f = S.lam (fun k -> S.app k (S.lam (fun x ->
    f (S.lam (fun k -> S.app k x)))))
  let app e1 e2 = S.lam (fun k -> 
    S.app e1 (S.lam (fun f ->
    S.app e2 (S.lam (fun v -> S.app (S.app f v) k)))))
  let fix f  = S.fix (fun self -> (f self))
end
\end{code}
The code literally establishes a mapping from CPS interpretations to
(direct) interpretations, that latter performed by 
the base interpreter |S|.

A reader may notice that the above module does not define the type
equation for |repr| and so it does not have the signature Symantics.
The reason for that is again the result type of |lam f|. Whereas
|int i| or |add e1 e2| have the (abbreviated) 
type |('c,..., (int -> 'w) -> 'w) S.repr|,
the result of |lam (fun x -> add x (int 1))| has the type
|('c,...((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr|. 
Hence, to write the type equation for |CPST.repr| we need a 
type case-discriminating
function similar to the type function |repr_pe| of
Section~\ref{S:PE-solution}. Alas, the function we need here is
similar but not identical to |repr_pe|. Thus we need to add 
another type variable to the |repr| type and modify the Symantics
signature accodringly. Again, the terms in previous Symantics module
stay unchanged, but the |repr| type equations in those structures have to
account for a new (phantom, in that case) type variable.

To save space, we show a simplified and less principal approach that
lets us use the module |CPST| as it is. Because the module does not
have the signature Symantics, we cannot apply the |EX| functor to it.
But we can write the tests nevertheless:
\begin{code}
module T = struct
 module M = CPST(C)
 open M
 let test1 () = app (lam (fun x -> x)) (bool true)
 let testpowfix () = ... (* the same as before *)
 let testpowfix7 = (* the same as before *)
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end;;
\end{code}
That is, we instantiate |CPST| with the desired base intrepreter (the
compiler |C| in the case above) so that we can use the result |M| to
interpret object level terms. The sample terms are \emph{exactly}
as we used before. We just copied them over. That copying step is the
price we pay for the simplified treatment. We shall encounter this
copying when discussing the self-interpreter in the next section. We
will see then that the copying step has a profound meaning of
`plugging the hole', which is one of the faces of polymorphism.

It is instructive to see the result of evaluating a few examples. With 
|CPST| instantiated with a straightforward compiler |C| as above, our
running test |T.test1 ()| gives
\begin{code}
.<fun x_5 -> ((fun x_2 -> 
        (x_2 (fun x_3 -> fun x_4 -> (x_4 x_3))))
        (fun x_6 -> ((fun x_1 -> (x_1 (true))) 
        (fun x_7 -> ((x_6 x_7) x_5)))))>.
\end{code}

which shows the naive CPS transform of the original code. The result is
quite unoptimal with several apparent $\beta$-redexes. We can reduce
them by instantiating |CPST| with the |P| evaluator instead. Our
running test gives then
\begin{code}
  P.repr = {P.st = Some <fun>; P.dy = .<fun x_5 -> (x_5 (true))>.}
\end{code}

@@

We can easily modify the CBV CPS transform to pass a piece of state
along with the continuation. This lets us support mutable state. We
can also use reference variables of the metalanguage to emulate
reference variables in the object language.
\oleg{Should we have an example in the code?}
